---
title: "portofolio opdracht 1.1"
author: "Arthur Timmermans"
date: '2022-04-29'
output: html_document
---

##lesson1 
A+B: 
```{r}
read_excel(here("raw/CE.LIQ.FLOW.062_Tidydata.xlsx"))
```

C: 
based upon the names, the expected data-types are: 
- Raw-data: nummeric
- compNames: character 
- compConcentration: nummeric 
In the imported table, Only the raw-data is nummeric, both the compNames and compConcentrations are nummeric. This would suggest that something went wrong during importation of the samples. 

D: 

```{r}
table1D <- read_excel(here("raw/CE.LIQ.FLOW.062_Tidydata.xlsx"))
ggplot(table1D, aes(x=compConcentration, y=RawData)) +geom_point() + theme(axis.text.x = element_text(angle = 90))
```

E: 
the data was not sorted based upon the actual size, becuase it is being read as character and therefor is litteraly sorted based upon the characters. 

F: 
```{r}
table1D$compConcentration <- as.numeric(table1D$compConcentration)
ggplot(table1D, aes(x=log10(compConcentration), y=RawData, color=expType)) +geom_point() + theme(axis.text.x = element_text(angle = 90)) + geom_jitter() + labs(title = "effect of compound concentration on amount of living C. elegans", y = "C. elegans count", x = "concentration of compound")
```

G+H: 
The positive control for this experiments is incubated in ethanol, the negative control for this experiment is incubated in s-medium 

I: 
- make a simpel dotted plot (already done) and look at how the data is roughly distributed. 
- check if the controls of the experiments have the expected result (positive control is positive, negative control is negative) 
- check if the data of each group is normally distributed (shapiro-wilk)
- compare the experimental group to the (positive) control group with the propper statistical tests. these would include: 
    - levene's test of equality of error variances
    - 2 sides unpaired t.test 
    

J: 


K: This brings the data in the plot relatively closer together, making it easier to compare data of different groups. 

