# loading, sorting and visualising data 

```{r setup1, echo=FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE)
library(tidyverse) 
library(readxl)
library(here)
library(ggplot2)
library(fs)
library(DT)
```

For both readability and data analysis is it important to keep data tidy. Especially when loading data in, it is important to check if the format is tidy for the data analysis that will be conducted. In this chapter, I will show an example of how I loaded in an excel document, selected the needed data and made the used tables tidy. The excel document contains data from an experiment conducted by  by J. Louter (INT/ILC).In this experiment, the influence of 3 substances on the growth of C. Elegans is tested. Further more, there is a positive and negative control. The positive control for this experiments is incubated in ethanol, the negative control for this experiment is incubated in s-medium. The raw data is shown in figure \@ref(fig:table1)

```{r table1, fig.cap= "table with raw data"}
datatable(read_excel(here::here("raw/CE.LIQ.FLOW.062_Tidydata.xlsx")), options = list(pageLength = 5)) 
```

based upon the names and values, as shown in figure \@ref(fig:table1), the expected data-types are: 

- Raw-data: numeric

- compNames: character 

- compConcentration: numeric 

In figure \@ref(fig:table1), Only the raw-data is numeric, both the compNames and compConcentrations are character strings. This would suggest that something went wrong during importation of the samples. This is also seen in figure \@ref(fig:graph1), where the data was not sorted based upon the actual size, because it is being read as character and therefor is literally sorted based upon the characters.

```{r graph1, fig.cap= "graph without changing data type"}
table1D <- read_excel(here("raw/CE.LIQ.FLOW.062_Tidydata.xlsx"))
ggplot(table1D, aes(x=compConcentration, y=RawData, color=compName, shape=expType)) +geom_point() + theme(axis.text.x = element_text(angle = 90))
```

 
When correcting the data type of the compound concentration back to numeric, the graph in figure \@ref(fig:graph2) can be created. 

```{r graph2, fig.cap= "graph with changing data type"}
table1D$compConcentration <- as.numeric(table1D$compConcentration)
ggplot(table1D, aes(x=log10(compConcentration), y=RawData, color=expType)) +geom_point() + theme(axis.text.x = element_text(angle = 90)) + geom_jitter() + labs(title = "effect of compound concentration on amount of living C. elegans", y = "C. elegans count", x = "concentration of compound")
```

In this graph, the data is still rather spread out. This can be corrected for by using the negative control as a scale of 1. This brings the data in the plot relatively closer together, making it easier to compare data of different groups. The graph would than look like shown in figure \@ref(fig:graph3) 

```{r graph3, fig.cap= "graph with normalized scale"}
table1clean <- data.frame(table1D$expType, table1D$RawData, table1D$compConcentration) #filtering for just exptype, raw data and compound concentration)
table1clean <-na.omit(table1clean) 

table1negative <- table1clean %>% filter(table1D.expType=="controlNegative") #filtering negative control 
average <- mean(table1negative$table1D.RawData) #adding average negative control to object 
table1clean$normalised <- table1clean$table1D.RawData/average #normalising with average negative control as 1. 

# making grahp: 
ggplot(table1clean, aes(x=log10(as.numeric(table1D.compConcentration)), y=normalised, color=table1D.expType)) +geom_point() + theme(axis.text.x = element_text(angle = 90)) + geom_jitter() + labs(title = "effect of compound concentration on amount of living C. elegans with normalised x axis", y = "C. elegans count", x = "concentration of compound")
```


## analysing the data: 
to further determine the (toxic) effect of the added substances, a dose-response analysis can be run.In this analysis, the following would be determined using a log-logistic model:

- estimate the maximal

- estimate the minimal 

- determine the IC50 concentration 

- determine the slope at IC50. 

the results of the different substances, can be compared to the positive and negative control, to determine if there is a toxic effect, and to what extend this toxic effect takes place. 

However, as analysis the entire data set is not the goal of this chapter, the further data analysis will not be conducted. But a r package that could be used, is the "dcr" package. 




